Predictive modeling using genomic annotations

### 2018-10-01 Manuscript/manuscript.Rmd
edits

In this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices at different resolutions for two cell lines, GM12878 and K562. We then performed a variety of machine learning algorithms on the pre-processed data to identify the genomic elements that were most predictive of TAD boundaries. Firstly, we found that the models performed on balanced data out performed those on unbalanced data. Secondly, it was determined that multiple iterations of random under-sampling was the balancing technique that yielded the best results. Thirdly, the inclusion of distances as predictors was found to improve model performance. Lastly, it was found that performing an inverse normal transformation of the distance predictors further improved model performance.

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 and K562 cell lines (Rao et. al).  The domain data was constructed from in situ Hi-C contact matrices at a 5kb and 10 kb resolution for the GM12878 and k562 cell lines respectively. Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, for GM12878, 5 kb genomic bins were used, while 10 kb bins were used for K562. Next, for the domain data provided by GM12878, the boundary coordinates were flanked on either side by 5 kb. For the K562, the boundary coordinates were flanked by 10 kb. In either case, a genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0). An illustration of the data is presented in Figure X.

@@

What normalization technique best improves model performance?

### VARIABLE SELECTION

Due to the large number of genomic features included in the model, different variable selection techniques were evaluated in order to improve computational speed and eliminate uninformative variables. There were three different wrapper methods that were used here, forward, backward, and stepwise selection. A null model consisting of just an intercept term and a full model, including all features, were specified for each selection technique. For all three, a logistic regression model was performed, and the AIC of the fitted model was used to establish which variables to include. That is, only a reduction in the model's AIC contributed to the inclusion of a feature in the model. For each technique, prior to model selection, balanced classes were iteratively created 100 times by random under-sampling from teh majority class. At each iteration the variables selected from each technique were recorded. The final set of predictors selected for each technique were those that appeared over 90% of the time accross all iterations. The union of the three sets of features from each of the techniques was obtained and used as the final feature space of the data (Figure X)

## FINAL MODEL

Once the pipeline produced the appropriate dataset, an elastic was used to assess which genomic features that were most predictive of the development of TAD boundaries. The algorithm was run iteratively over 100 random under-sampled majority classes to balance the data, each time producing performance metrics and variable importances. At the end of the 100 iterations, row-wise means were performed to aggregate the model information. We then compared the elastic-net model performed on the processed data to both a multiple logistic regression model and a LASSO regularized model with and without the use of distance predictors and with or without processed data.

#####################

>>>>

In this study we have developed a novel data pre-processing pipeline that addresses the irregularities in the TAD boundary data provided by contact matrices for the GM12878 cell line. We then performed a variety of machine learning algorithms on the pre-processed data to identify the genomic elements that were most predictive of TAD boundaries. Firstly, we found that the models performed on balanced data out performed those on unbalanced data. Secondly, it was determined that multiple iterations of random under-sampling was the balancing technique that yielded the best results. Thirdly, the inclusion of distances as predictors was found to improve model performance. Lastly, it was found that performing an inverse normal transformation of the distance predictors further improved model performance.

@@

Publicly available topologically associating domain data was obtained from GEO with accession GSE53525 for the GM12878 cell line (Rao et. al).  The domain data was constructed from in situ Hi-C contact matrices at a 5kb resolution. Identified TAD boundaries were represented by their genomic coordinates (hg19 human genome assembly), including chromosome, start and end coordinates. The start and end coordinates were concatenated, sorted, and unique coordinates representing the borders demarcating TADs were obtained.

The establishment of the data for modeling first consisted of binning the genome into equal sized intervals. This was performed at the same resolution as the respective cell line. That is, 5 kb genomic bins were used. Next, the boundary coordinates were flanked on either side by 5 kb. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing an boundary (Y=0). An illustration of the data is presented in Figure X.

@@

What normalization technique performed on the distance predictors best improves model performance? Here, we considered three normalization techniques. The first was a simple $log_{2}$ transformation, in which one was added in the case of distances of 0. The second was an inverse hyperbolic sine tranformation given by $insert function here$ which has the benefit of handling values of 0. Lastly, an inverse normal tranformation was considered. The inverse normal transformation is a way of transforming the sample distribution of a continuous variable to make it appear more normally distributed by matching the normalized rank of the trait to a quantile in a normal distribution using the following formula $insert formula here$. Multiple logistic regression models were performed using data with each respective transformation and compared. It was found that the model with inverse normally transformed distance predictors not only out-performed the other transformation but also out-performed a model with no transformation at all. So far the only classification algorithm considered was a multiple logistic regression. However, given the large number possible functional genomic elements that can be considered, this model can be computationally inefficient.

### Question 5: Classification

Which classification algorithm most accurately and efficiently identifies the genomic elements associated with TAD boundaries?
